The Talking Heads Experiment

### 2015-04-16 chap9/chap9.tex
hyphenation, orphans, widows and some other layout updated

\subsection{The Exhibition}

The N01SE exhibition\is{N01SE exhibition} was about information and transformation. 
Various locations in Cambridge (UK) participated: The Kettle's Yard university gallery, 
the Cambridge Whipple Museum of the History of Science, the Cambridge University Museum of 
Archaeology and Anthropology and the 
Fritzwilliam Museum. The installations ran from 22 January until 26 March 2000. There was also a site 
in London at the Wellcome Trust Two10 Gallery in Euston Road, which 
ran from 28 January to 1 May 2000. Apart from the usual press coverage for art exhibitions, articles appeared in
Science\footnote{http://www.firstpulseprojects.com/sciencerev.html}
and the Lancet \footnote{http://www.garnettmckeen.net/lancet.html}
showing that the exhibition resonated also in the scientific press.

@@

The exhibition was anounced in the following way by its curators: 
\begin{quotation}
{\itshape A multi-site multimedia exhibition in Cambridge with ``realtime" links to London, organised around three key themes in ``digitality'':

@@

N01SE celebrates the essential excess from which information is drawn. It probes many different ways of seeing and 
being in the world. Chances are your own sense of order is already someone elses N01SE.}
\end{quotation}
%\begin{figure}[htbp]
%  \centerline{\includegraphics[width=.95\textwidth]{chap9/figs/latour.pdf}}
%\caption{\label{latour} 
%Luc Steels Discusses The N01SE exhibition with Bruno Latour who intensively followed the experiments already from the first series at the Laboratorium exhibition in Antwerp. His comments significantly improved the initial designs.}
%\end{figure}

@@

The Kettle's Yard Gallery is associated with Cambridge University. It acted as the main site of the N01SE exhibition 
and showed a variety of historical artefacts (including the brain of Babbage and the original DNA structure built by 
Watson and Crick) together with new artistic works. The catalogueue published by the Kettle's Yard 
Gallery \footnote{www.kettlesyard.co.uk/noise} featured articles by Brian Smith, Umberto Eco, Bruno Latour, 
Bruce Sterling, Luc Steels, Peter Weibel and others. Below is my text as it appeared in the catalogue: 
\begin{quotation}
``Meanings are not a priori Platonic entities independent of language; meanings are the result of embodied interactions with the world, obtained via the role words play in verbal interactions called ``language games''. The Talking Heads Experiment explores one kind of language game: a guessing game played by two robotic agents about the scene directly in front of them. One agent acts as Speaker and attempts to draw the attention of the Hearer to some object by transmitting a verbal description of it; the Speaker succeeds if the Hearer correctly guesses which object is ``meant''.

@@

To express conceptualisations, agents need a lexicon relating words to meanings. It must function 
bi-directionally (words-to-meanings and meanings-to- words); it must also store synonyms (more than one word for the same meaning) and ambiguities (more than one meaning for the same word). The agents have not been given a lexicon; they must acquire their own common lexicon as a bi-product of the game.

@@

A group of University of Hull students (with names like {\it Trash}, {\it yeah8a8y}, {\it schedski}) 
became very much involved. They 
communicated through the Forum using often slang and sexist language and making a surprising 
number of spelling errors and grammatical mistakes. The group had been 
trying to impose their own language by very actively teaching their agents and sending them to the same site, namely 
Paris, which they had noticed had the most stable operating conditions. This worked (as indeed it was supposed to) 
at which point they decided to do the same with the London site although the camera 
alignment was too unreliable to allow the evolution of a stable language. 
\newline

@@

{\bfshape 2000-02-25 21:55:49 Trash}\\
{\itshape YELLOW MK 1 ESCORT}\\
Hey mate nice seeing you around paris !!!! Enjoy the company especially since you know most of our words AND get the Biatches right... you can have my sister anyday... so then mate see ya round keep driving the crap car.	
Have noticed the absence of fine ass girls in here am i wrong.... hope soooo.
\end{mail}

@@

{\bfshape 2000-02-25 22:02:36 LondonCalling}\\
{\itshape CALIBRATION OF CAMERAS London}\\
"Can London PLEASE calibrate their cameras???? PLEASE!!!!!\\

@@

{\bfshape 2000-02-26 16:52:32 Yeah8a8y}\\
{\itshape RE: CALIBRATION OF CAMERAS London}\\
very good.... Absolutely no chance of Domination if the cameras arn't calibrated... SORT IT!!\\

@@

{\bfshape 2000-03-02 15:46:50} phlox\\
{\itshape RE: CALIBRATION OF CAMERAS}\\
i manage the two10 gallery in which the talking heads display is currently installed. would that it were so simple to sort it and keep the cameras correctly calibrated. all equipment was installed by the brussels group and whenever the cameras need calibrating - for whatever reason - it means one of the brussels boffins has to pop over to fix it. they've been over once since the exhibition opened and decided to move the cameras forward 2 feet. not really an exact science is it? the grand master - luc steels - is coming to the gallery tomorrow, so perhaps he will enlighten me as to why the cameras seem to need constant calibration, and i'll let you know. cheers.\\

@@

{\bfshape 2000-03-02 20:04:52 fish}\\
{\itshape RE: CALIBRATION OF CAMERAS London}\\
that's as maybe, phlox, but i'm guessing you secretly come into the exhibition and kick the cameras when 
you're in a bad mood, like say when someone has been horrid to you.\\

@@

{\bfshape 2000-03-03 20:08:36 schedski}\\
{\itshape RE: CALIBRATION OF CAMERAS London}\\
wassup fish?\\

@@

{\bfshape 2000-03-03 23:06:43 fish}\\
{\itshape RE: CALIBRATION OF CAMERAS London}\\
come on, admit it, you guys there spend all day stomping around your exhibits, breaking the displays, unplugging the monitors and generally mucking things up. i'm surprised london's still got two cameras even working and they haven't already been flogged down some boozer for a monkey and a couple of jellied eels\\

@@

{\bfshape 2000-03-07 13:03:46 A Londoner}\\
{\itshape RE: CALIBRATION OF CAMERAS London}\\
Well, if you people will confuse the poor things with your outrageous diction, they might know which way to point... One of them was practically in tears the other day as I watched it shaking its head in disbelief at the language it was meant to repeat. What ever happened to `Adam's a cool bloke'?
\end{mail}

Meanwhile the Hull group is pursuing and explaining their particular experiment that they carry out within the 
Talking Heads framework. The experiment itself is interesting and shows that they understand what is 
going on and are creative. Also the overall dynamics is working. The group realises that a language can form 
on its own or that it can be influenced by humans teaching agents their own language. 
At the same time, there is already one user (called Cheesy) who is pushing to introduce
dirty words to the agents. He does not belong to the Hull group but is most probably somebody from the 
Wellcome Gallery already trying to put the N01SE exhibition in a bad light. \\

@@

{\bfshape 2000-02-27 19:39:10 Trash}\\
{\itshape RE: teaching / TRIBES}\\
You mean by not playing god ???? and letting them get on with it at one site?? .... well yes but the server would have to be locked down by that tribe otherwise you would get outside inflences teaching them stupid like gumble which meanbs everything... it would probably take just 100-300 games each agent to develop a usable language (Aslong as the server has no agents from outside the  tribe in there)\\
By teaching them myslef and Yeahbaby got usable language going in  less than 2 days!!!!!!! Just see the success of most of our words in the lexicon top 20!!\\
God us 2 are good!!\\

@@

\rule{0.8\textwidth}{.4pt}
\end{mail}

To deal with the problem of agent congestion, software issues, and camera misalignment at the London 
and Cambridge sites, maintenance was carried out which lead to a temporary unavailability of the experiment. 
Surprisingly, the reaction of the Hull group was entirely negative as they saw it as a 
way to prevent them from gaining or keeping control of the language used by the agents, even though 
Angus McIntyre clearly and patiently explained why maintenance was necessary. Also, Denna Jones ({\it Phlox}) took 
it personally as she did not realise that the calibration errors were due to vibrations caused by
traffic of the London Underground near and under the Wellcome Gallery.  \\

@@

\rule{0.8\textwidth}{.4pt}
\end{mail}

@@

\rule{0.8\textwidth}{.4pt}
\end{mail}

The postings on the Forum keep going in crescendo for a while until the exhibition ends March 2000.

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.65\textwidth]{chap9/figs/th4.jpg}}
\caption{\label{fig:palais}Talking Heads installation at the Palais de la D\'{e}couverte. It featured new fancy structures that 
made the installation more attractive visually. Daily explanations were given to visitors 
by the staff of the Palais de la D\'{e}couverte.} 
\end{figure}

@@

1. The {\itshape European Conference on Artificial Life} at the EPFL in Lausanne (Switzerland) in September 1999. 
Papers on computational and robotic 
models of language evolution were (and still are) greeted with hostility at linguistics conferences (even 
conferences on computational linguistics) and they are still routinely rejected for linguistics journals, as 
being irrelevant for understanding more about human language. 
However the Artificial Life conferences and journals welcomed the approach from the beginning. This is 
not surprising because agent-base modeling is one of the main tools used in that field and an evolutionary 
stance is seen as obvious to biologists. The Lausanne conference was organised in line with earlier 
conferences on artificial life, showing work on life-like robots, computer simulations, and new chemically based 
forms of life. It also featured a series of live demos. The portable demonstration of the Talking Heads was part 
of these demonstrations, set up to accompany a presentation at 
the conference by myself and Fr\'{e}deric Kaplan\footnote{
The paper published for this conference is \cite{Steels:1999}.}
It was linked in with the ongoing experiment during Laboratorium in Antwerp.

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.90\textwidth]{chap9/figs/Lausanne.pdf}}
\caption{\label{fig:lausanne}Live installation and demonstration of the Talking Heads experiment at the European Conference on Artificial 
Life in 1999.}
\end{figure}

2. The {\itshape Neuer Aachener Kunstverein} in Aachen (Germany) in collaboration with the 
RWTH (the technical university of Aachen) organised a general exhibition on modelling called 
Modell-Modell. Within this framework, artist Anne-Mie van Kerckhoven invited me to cooperate in a laboratory on 
language and colour called ``Cyberlabor Chromosophy''. As part of this laboratory, the portable Talking Heads 
experiment was installed and ran from 5 May to 16 June 2000. It featured not only the portable Talking Heads, 
which was demonstrated live, but also posters, talks about the project and additional art works.

3. The portable Talking Heads was also featured in an exhibition at the {\itshape Ludo Mich Gallery} in Antwerp. This 
exhibition focused on colour again and showed several other pieces related to colour and colour perception. 
It was accompanied by a very well attended gallery talk.

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.80\textwidth]{chap9/figs/mich-gallery.pdf}}
\caption{\label{fig:mich}Live installation at the gallery of Ludo Mich in Antwerp. There was no white board but different 
small pancartes with possible scenes (see in the middle of the picture). The game was displayed much larger against
the back wall.}
\end{figure}

>>>>

\subsection{The Exhibition} 
\enlargethispage{2\baselineskip}
The N01SE exhibition\is{N01SE exhibition} was about information and transformation. 
Various locations in Cambridge (UK) participated: The Kettle's Yard university gallery, 
the Cambridge Whipple Museum of the History of Science, the Cambridge University Museum of 
Archaeology and Anthropology and the 
Fritzwilliam Museum. The installations ran from 22 January until 26 March 2000. There was also a site 
in London at the Wellcome Trust Two10 Gallery in Euston Road, which 
ran from 28 January to 1 May 2000. Apart from the usual press coverage for art exhibitions, articles appeared in Science\footnote{http://www.firstpulseprojects.com/sciencerev.html} and the Lancet\footnote{http://www.garnettmckeen.net/lancet.html} showing that the exhibition resonated also in the scientific press.

@@

The exhibition was anounced in the following way by its curators: 
\begin{quotation}
A multi-site multimedia exhibition in Cambridge with ``realtime" links to London, organised around three key themes in ``digitality'':

@@

N01SE celebrates the essential excess from which information is drawn. It probes many different ways of seeing and 
being in the world. Chances are your own sense of order is already someone elses N01SE.\todo{quotations are not set in italics, see GSR}
\end{quotation}
%\begin{figure}[htbp]
%  \centerline{\includegraphics[width=.95\textwidth]{chap9/figs/latour.pdf}}
%\caption{\label{latour} 
%Luc Steels Discusses The N01SE exhibition with Bruno Latour who intensively followed the experiments already from the first series at the Laboratorium exhibition in Antwerp. His comments significantly improved the initial designs.}
%\end{figure}

@@

``The Kettle's Yard Gallery is associated with Cambridge University. It acted as the main site of the N01SE exhibition 
and showed a variety of historical artefacts (including the brain of Babbage and the original DNA structure built by 
Watson and Crick) together with new artistic works. The catalogueue published by the Kettle's Yard 
Gallery \footnote{www.kettlesyard.co.uk/noise} featured articles by Brian Smith, Umberto Eco, Bruno Latour, 
Bruce Sterling, Luc Steels, Peter Weibel and others. Below is my text as it appeared in the catalogue: 
\begin{quotation}
Meanings are not a priori Platonic entities independent of language; meanings are the result of embodied interactions with the world, obtained via the role words play in verbal interactions called ``language games''. The Talking Heads Experiment explores one kind of language game: a guessing game played by two robotic agents about the scene directly in front of them. One agent acts as Speaker and attempts to draw the attention of the Hearer to some object by transmitting a verbal description of it; the Speaker succeeds if the Hearer correctly guesses which object is ``meant''.

@@

To express conceptualisations, agents need a lexicon relating words to meanings. It must function 
bi-directionally (words-to-meanings and meanings-to-words); it must also store synonyms (more than one word for the same meaning) and ambiguities (more than one meaning for the same word). The agents have not been given a lexicon; they must acquire their own common lexicon as a bi-product of the game.

@@

A group of University of Hull students (with names like {\it Trash}, {\it yeah8a8y}, {\it schedski}) 
became very much involved. They 
communicated through the Forum using often slang and sexist language and making a surprising 
number of spelling errors and grammatical mistakes. The group had been 
trying to impose their own language by very actively teaching their agents and sending them to the same site, namely 
Paris, which they had noticed had the most stable operating conditions. This worked (as indeed it was supposed to) 
at which point they decided to do the same with the London site although the camera 
alignment was too unreliable to allow the evolution of a stable language.

@@

{\bfshape 2000-02-25 21:55:49 Trash}\\
{\itshape \textsc{yellow} MK 1 ESCORT}\\
Hey mate nice seeing you around paris !!!! Enjoy the company especially since you know most of our words AND get the Biatches right... you can have my sister anyday... so then mate see ya round keep driving the crap car.	
Have noticed the absence of fine ass girls in here am i wrong.... hope soooo.
\end{mail}

@@

{\bfshape 2000-02-25 22:02:36 LondonCalling}\\
{\itshape CALIB\textsc{ratio}N OF CAMERAS London}\\
"Can London PLEASE calibrate their cameras???? PLEASE!!!!!\\

@@

{\bfshape 2000-02-26 16:52:32 Yeah8a8y}\\
{\itshape RE: CALIB\textsc{ratio}N OF CAMERAS London}\\
very good.... Absolutely no chance of Domination if the cameras arn't calibrated... SORT IT!!\\

@@

{\bfshape 2000-03-02 15:46:50} phlox\\
{\itshape RE: CALIB\textsc{ratio}N OF CAMERAS}\\
i manage the two10 gallery in which the talking heads display is currently installed. would that it were so simple to sort it and keep the cameras correctly calibrated. all equipment was installed by the brussels group and whenever the cameras need calibrating - for whatever reason - it means one of the brussels boffins has to pop over to fix it. they've been over once since the exhibition opened and decided to move the cameras forward 2 feet. not really an exact science is it? the grand master - luc steels - is coming to the gallery tomorrow, so perhaps he will enlighten me as to why the cameras seem to need constant calibration, and i'll let you know. cheers.\\

@@

{\bfshape 2000-03-02 20:04:52 fish}\\
{\itshape RE: CALIB\textsc{ratio}N OF CAMERAS London}\\
that's as maybe, phlox, but i'm guessing you secretly come into the exhibition and kick the cameras when 
you're in a bad mood, like say when someone has been horrid to you.\\

@@

{\bfshape 2000-03-03 20:08:36 schedski}\\
{\itshape RE: CALIB\textsc{ratio}N OF CAMERAS London}\\
wassup fish?\\

@@

{\bfshape 2000-03-03 23:06:43 fish}\\
{\itshape RE: CALIB\textsc{ratio}N OF CAMERAS London}\\
come on, admit it, you guys there spend all day stomping around your exhibits, breaking the displays, unplugging the monitors and generally mucking things up. i'm surprised london's still got two cameras even working and they haven't already been flogged down some boozer for a monkey and a couple of jellied eels\\

@@

{\bfshape 2000-03-07 13:03:46 A Londoner}\\
{\itshape RE: CALIB\textsc{ratio}N OF CAMERAS London}\\
Well, if you people will confuse the poor things with your outrageous diction, they might know which way to point... One of them was practically in tears the other day as I watched it shaking its head in disbelief at the language it was meant to repeat. What ever happened to `Adam's a cool bloke'?
\end{mail}

Meanwhile the Hull group is pursuing and explaining their particular experiment that they carry out within the 
Talking Heads framework. The experiment itself is interesting and shows that they understand what is 
going on and are creative. Also the overall dynamics is working. The group realises that a language can form 
on its own or that it can be influenced by humans teaching agents their own language. 
At the same time, there is already one user (called Cheesy) who is pushing to introduce
dirty words to the agents. He does not belong to the Hull group but is most probably somebody from the 
Wellcome Gallery already trying to put the N01SE exhibition in a bad light.

@@

{\bfshape 2000-02-27 19:39:10 Trash}\\
{\itshape RE: teaching / TRIBES}\\
You mean by not playing god ???? and letting them get on with it at one site?? .... well yes but the server would have to be locked down by that tribe otherwise you would get outside inflences teaching them stupid like gumble which meanbs everything... it would probably take just 100-300 games each agent to develop a usable language (Aslong as the server has no agents from outside the  tribe in there)\\
By teaching them myslef and Yeahbaby got usable language going in  less than 2 days!!!!!!! Just see the success of most of our words in the lexicon top 20!!\\ \enlargethispage{2\baselineskip}
God us 2 are good!!\\

@@

% \rule{0.8\textwidth}{.4pt}
\end{mail}
\todo{removed one \textbackslash rule from the last entry. Also, why is \textbackslash rule set to .8 of textwidth?!}
\enlargethispage{1\baselineskip}
To deal with the problem of agent congestion, software issues, and camera misalignment at the London 
and Cambridge sites, maintenance was carried out which lead to a temporary unavailability of the experiment. 
Surprisingly, the reaction of the Hull group was entirely negative as they saw it as a 
way to prevent them from gaining or keeping control of the language used by the agents, even though 
Angus McIntyre clearly and patiently explained why maintenance was necessary. Also, Denna Jones ({\it Phlox}) took 
it personally as she did not realise that the calibration errors were due to vibrations caused by
traffic of the London Underground near and under the Wellcome Gallery.

@@

\end{mail}

@@

\end{mail}
\clearpage
The postings on the Forum keep going in crescendo for a while until the exhibition ends March 2000.

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=\textwidth]{chap9/figs/th4.jpg}}
\caption{\label{fig:palais}Talking Heads installation at the Palais de la D\'{e}couverte. It featured new fancy structures that 
made the installation more attractive visually. Daily explanations were given to visitors 
by the staff of the Palais de la D\'{e}couverte.} 
\end{figure}

@@

\begin{enumerate}
\item The {\itshape European Conference on Artificial Life} at the EPFL in Lausanne (Switzerland) in September 1999. 
Papers on computational and robotic 
models of language evolution were (and still are) greeted with hostility at linguistics conferences (even 
conferences on computational linguistics) and they are still routinely rejected for linguistics journals, as 
being irrelevant for understanding more about human language. 
However the Artificial Life conferences and journals welcomed the approach from the beginning. This is 
not surprising because agent-base modeling is one of the main tools used in that field and an evolutionary 
stance is seen as obvious to biologists. The Lausanne conference was organised in line with earlier 
conferences on artificial life, showing work on life-like robots, computer simulations, and new chemically based 
forms of life. It also featured a series of live demos. The portable demonstration of the Talking Heads was part 
of these demonstrations, set up to accompany a presentation at 
the conference by myself and Fr\'{e}deric Kaplan\footnote{
The paper published for this conference is \cite{Steels:1999}.}
It was linked in with the ongoing experiment during Laboratorium in Antwerp. 
\item The {\itshape Neuer Aachener Kunstverein} in Aachen (Germany) in collaboration with the 
RWTH (the technical university of Aachen) organised a general exhibition on modelling called 
Modell-Modell. Within this framework, artist Anne-Mie van Kerckhoven invited me to cooperate in a laboratory on 
language and colour called ``Cyberlabor Chromosophy''. As part of this laboratory, the portable Talking Heads 
experiment was installed and ran from 5 May to 16 June 2000. It featured not only the portable Talking Heads, 
which was demonstrated live, but also posters, talks about the project and additional art works. 
\item The portable Talking Heads was also featured in an exhibition at the {\itshape Ludo Mich Gallery} in Antwerp. This 
exhibition focused on colour again and showed several other pieces related to colour and colour perception. 
It was accompanied by a very well attended gallery talk. 
\end{enumerate}
\begin{figure}[htbp]
  \centerline{\includegraphics[width=.90\textwidth]{chap9/figs/Lausanne.pdf}}
\caption{\label{fig:lausanne}Live installation and demonstration of the Talking Heads experiment at the European Conference on Artificial 
Life in 1999.}
\end{figure}
\begin{figure}[htbp]
  \centerline{\includegraphics[width=.80\textwidth]{chap9/figs/mich-gallery.pdf}}
\caption{\label{fig:mich}Live installation at the gallery of Ludo Mich in Antwerp. There was no white board but different 
small pancartes with possible scenes (see in the middle of the picture). The game was displayed much larger against
the back wall.}
\end{figure}
